# **R for Data Science - Whole Game**
This document covers my journey through the R for Data Science content hosted at https://r4ds.hadley.nz/

# 1 Data Visualization

## First Steps

```{r}
library(tidyverse) # essential for working with tidy data

library(palmerpenguins) # contains datasetYes

library(ggthemes) # themes for ggplot?

```

### Terms
- Variable: Quantity, quality, or property that you can measure
- Value: The state of a variable when you measure it
- Observation: A set of measurements made under similar conditions
- Tablular data: Set of values, each associated with a variable and an observation
  - Considered **tidy** if each value is placed in its own cell, each variable it its own column, and each observation in its own row
- Tibble: An extended dataframe with additional information

```{r}

# Looking at summaries of datasets
glimpse(penguins)
View(penguins) # interactive viewer in RStudio
```

### ggplot Overview
**aes (Aesthetics)**: Aesthetics of the graph
**geom (Geometrics)**: Geometrical object to represent data


### Goal
Create a visualization to display the relationship between flipper lengths and body masses of the penguin species within the penguins dataset. 


```{r}
ggplot(
  data=penguins,
  mapping = aes(x=flipper_length_mm, y=body_mass_g) # aes is aesthetics
) + 
  geom_point(mapping = aes(color=species, shape=species)) + # geometrical object to represent data on plot, do species coloring here as to not change ggplot aes
  geom_smooth(method='lm') + # add best fit line
  labs(
    title="Body Mass and Flipper Length",
    subtitle="Dimensions for Species of Penguins",
    x="Flipper length (mm)",
    y="Body mass (g)",
    color="Species",
    shape="Species"
  ) +
  scale_color_colorblind() # from ggthemes
```

### Exercises

1. How many rows, columns in penguins?
From the environment: 344 rows, 8 columns
```{r}
df <- penguins
```
2. What does bill_depth_mm describe? 
It is a number denoting bill depth in millimeters
```{r}
?penguins
```
3. Make scatterplot of bill_depth_mm vs. bill_length_mm (x-axis). Describe relationship.
Appears to be 2 or 3 clusters. Higher bill depth could have shorter bill length but not in all cases. Suspect species are the clusters. By adding color and shape to points this is confirmed. 
```{r}
ggplot(
  data=penguins,
  mapping=(aes(x=bill_length_mm, y=bill_depth_mm))
) +
  geom_point(mapping=aes(color=species, shape=species))
```
4. What happens if you make a scatterplot of species vs. bill_depth_mm? What might be a better choice of geom?
When making a scatterplot of this data it appears as lines of points which are difficult to interpret. A boxplot would best show this data as we could see the general distribution of bill depth among the species. 
```{r}
ggplot(
  data=penguins,
  mapping=aes(x=species, y=bill_depth_mm)
) +
  geom_boxplot()

```
5. Why does the following give an error and how would you fix it?
The error occurs because no x and y values are specified in aesthetics so a geometric object cannot be plotted. Specifically, geom_point() requies x and y to be set in the aesthetics of layer 1, which is ggplot. To fix this, add the variables which are desired to plot. 
```{r}
ggplot(data=penguins) +
  geom_point()
```

6. What does the na.rm argument do in geom_point()? What is the default value of the argument? Create a scatterplot where you successfully use this argument set to TRUE.
The na.rm argument determines if missing values are removed with a warning. It is FALSE, by default, but if set to TRUE then missing values are silently removed. This argument can help with hiding the warning that 2 values are removed when plotting previous graphs. 
```{r}
#?geom_point()
ggplot(
  data=penguins,
  mapping=aes(x=bill_length_mm, y=flipper_length_mm)
) +
  geom_point(na.rm=TRUE)

```
7. Add the following caption to the plot in Q6: "Data comes from the palmerpenguins package."
```{r}
ggplot(
  data=penguins,
  mapping=aes(x=bill_length_mm, y=flipper_length_mm)
) + 
  geom_point(na.rm=TRUE) +
  labs(caption="Data comes from the palmerpenguins package.")

```
8. Recreate the visualization. What aesthetic should bill_depth_mm be mapped to? And should it be mapped at the global level or at the geom level?
The color aesthetic should be mapped to bill_depth_mm at the geom level so that the mapping does not effect other upper layer additions to the chart, such as geom_smooth().
```{r}
ggplot(
  data=penguins,
  mapping=aes(x=flipper_length_mm, y=body_mass_g)
) +
  geom_point(mapping=aes(color=bill_depth_mm)) +
  geom_smooth() # draws a curved line without 'lm'

```
9. Run this code in your head and predict what the output will look like. Then run the code and check predictions. 
Prediction: Larger body mass will correlate with longer flipper length (bigger animals). Species will have their own relationships between the variables. The points will form 3 clusters. There will be 3 distinct smooth lines displaying the relationship in each species.
Check: Did not reflect on island being the color, Biscoe seems to hold mainly penguins with longer flippers and higher body mass. Overall seems to be a positive relation between the factors, as predicted. Species clusters would require additional aesthetics, such as shape. Adding this confirms predictions roughly, though they are hard to identify clearly.
```{r}
ggplot(
  data=penguins,
  mapping=aes(x=flipper_length_mm, y=body_mass_g, color=island)
) +
  geom_point(mapping=aes(shape=species)) +
  geom_smooth(se=FALSE)
```
10. Will these two graphs look different? Why/why not?
Prediction: They will not look different, as they both have the same aesthetic information. Nevertheless, the second option is less efficient and does not take advantage of the layered approach to ggplot graphing. 
```{r}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point() +
  geom_smooth()

ggplot() +
  geom_point(
    data = penguins,
    mapping = aes(x = flipper_length_mm, y = body_mass_g)
  ) +
  geom_smooth(
    data = penguins,
    mapping = aes(x = flipper_length_mm, y = body_mass_g)
  )
```
## 1.3 ggplot2 calls
Improving conciseness
```{r}
# v1
ggplot(
  data=penguins,
  mapping=aes(x=flipper_length_mm, y=body_mass_g)
) +
  geom_point()

# v2
ggplot(penguins, aes(x=flipper_length_mm, y=body_mass_g)) +
  geom_point()

# v3
penguins |>
  ggplot(aes(x=flipper_length_mm, y=body_mass_g)) +
  geom_point()
```

## 1.4 Visualizing distributions
### Categorical variable
Can only take one of a small set of values. Ex. Species, island
```{r}
ggplot(penguins, aes(x=species)) +
  geom_bar()

ggplot(penguins, aes(x=island)) +
  geom_bar()
```
Order based on frequency
```{r}
ggplot(penguins, aes(x=fct_infreq(species))) +
  geom_bar()
```

### Numerical variable
Can take a wide range of numerical values. Can be discrete or continuous.

Histogram divides x-axis into equally spaced bins and shows frequencies. Play with the bin value to best represent the data. 
```{r}
ggplot(penguins, aes(x=body_mass_g)) +
  geom_histogram(binwidth=200)
```
Alternative to histogram is a density plot. It displays frequency with a smooth line.
```{r}
ggplot(penguins, aes(x=body_mass_g)) +
  geom_density()
```
### Exercises
1. Make a bar plot of species of penguins, assign species to y. How is this plot different?
This plot is also showing the counts of species. With the variable being assigned to y the plot displays horizontally instead of vertically. 
```{r}
ggplot(penguins, aes(y=species)) +
  geom_bar()
```
2. How are the following plots different? Which aesthetic, color or fill, is more useful for changing the color of bars?
The plots use color vs. fill in the bar charts. Color is meant to draw lines and strokes, while fill is used to color inside areas of geoms; from this documentation one can predict fill is the better option for bar charts. Executing the code confirms this, as color only outlines the bars.
```{r}
ggplot(penguins, aes(x = species)) +
  geom_bar(color = "red")

ggplot(penguins, aes(x = species)) +
  geom_bar(fill = "red")
```
3. What does the bins argument in geom_histogram() do?
The bins argument allows one to hardcode the number of bins to display. This will be overridden if binwidths is set (bin count determined dynamically). The default bin count is 30. 

4. Make a histogram of the carat variable in the diamonds dataset. Experiment with different binwidths. What binwidth reveals the most interesting pattern?
Small binwidth shows the frequency around 1 carat. Lowing the binwidth into small decimal points reveals many peaks of carats all around the 0 to 2 carat range.
```{r}
ggplot(diamonds, aes(x=carat)) +
  geom_histogram(binwidth=0.005)
```


# 2 Workflow: basics


